Verified Nanopasses for Compiling Conditionals

### 2025-06-20 paper.tex
minor edit

%% \author{Lars Th{\o}rv{\"a}ld}
%% \affiliation{%
%%   \institution{The Th{\o}rv{\"a}ld Group}
%%   \city{Hekla}
%%   \country{Iceland}}
%% \email{larst@affiliation.org}

%% \author{Valerie B\'eranger}
%% \affiliation{%
%%   \institution{Inria Paris-Rocquencourt}
%%   \city{Rocquencourt}
%%   \country{France}
%% }

%% \author{Aparna Patel}
%% \affiliation{%
%%  \institution{Rajiv Gandhi University}
%%  \city{Doimukh}
%%  \state{Arunachal Pradesh}
%%  \country{India}}

%% \author{Huifen Chan}
%% \affiliation{%
%%   \institution{Tsinghua University}
%%   \city{Haidian Qu}
%%   \state{Beijing Shi}
%%   \country{China}}

%% \author{Charles Palmer}
%% \affiliation{%
%%   \institution{Palmer Research Laboratories}
%%   \city{San Antonio}
%%   \state{Texas}
%%   \country{USA}}
%% \email{cpalmer@prl.com}

%% \author{John Smith}
%% \affiliation{%
%%   \institution{The Th{\o}rv{\"a}ld Group}
%%   \city{Hekla}
%%   \country{Iceland}}
%% \email{jsmith@affiliation.org}

%% \author{Julius P. Kumquat}
%% \affiliation{%
%%   \institution{The Kumquat Consortium}
%%   \city{New York}
%%   \country{USA}}
%% \email{jpkumquat@consortium.net}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
We present a proof of correctness in Agda for four nanopasses that
translate a source language with let binding, integer arithmetic,
conditional expressions and Booleans into an x86-flavored register
transfer language. The most interesting of these four nanopasses is a
translation of conditional expressions into goto-style control flow
that uses the continuation-oriented approach of Olivier Danvy's
one-pass transformation into monadic normal form (2003).
\end{abstract}

@@

The compilation of Boolean expressions and conditionals is a classic
topic in the compiler literature, going back to an algorithm of
\citet{Aho:1986qf} (presented in the form of an attribute grammar)
that translates Boolean expressions and conditionals into gotos. The
compilation of Boolean expressions and conditionals often occurs at an
interesting stage in the compilation pipeline, where the intermediate
language changes from an abstract syntax \emph{tree} into a control
flow \emph{graph}, or equivalently, an intermediate language with
\texttt{goto}. Further, when targeting an assembly language such as
Intel's x86, the naive approach to compiling boolean expressions and
conditionals produces embarassingly inefficient code. Similar issues
arise in the compilation of languages such as Haskell, where the
case-of-case transformation is employed to optimize dispatching on
algebraic datatypes~\citep{PEYTONJONES19983}.

@@

My first exposure to these ideas were in Kent Dybvig's compiler
course~\citep{Dybvig:2010aa} where he taught the techniques used in
the Chez Scheme compiler~\citep{Dybvig:2006aa}.  One of those
techniques is called destintation-driven code
generation~\cite{Dybvig:1990aa}, which adds an extra parameter to the
compilation function to indicate how the context will use the result
value of an expression. The context may (1) ignore the result (and
only care about the expression's side effects), (2) branch based on
interpreting its result as a boolean, (3) assign the result to a
variable, or (4) output the result. These options are encoded as a
disjoint union.  Relating this to Olivier's transformation functions,
the $\mathcal{E}$ function is for contexts that need the actual value
and the $\mathcal{B}$ function is for contexts that branch on the
expression's result. Of course, a pair of functions is isomorphic to a
function on a disjoint union.
\[
(A → C) × (B → C) ≅ (A + B) → C
\]

@@

Towards developing a course on verified compilation, I am interested
in adapting verification techniques from CompCert to the compiler
presented in the \emph{Essentials of Compilation}
textbook~\citep{Siek:2023tr,Siek:2023ue}. That book takes an
incremental approach to teaching compilers, that is, it guides the
students through the creation of multiple compilers for a sequence of
languages, each of which adds a language feature to the previous
langauge. For the language with Boolean expressions and conditionals,
the Explicate Control pass of the compiler uses the one-function per
context approach of Danvy to translate into an x86-flavored register
transfer language.  \emph{Essentials of Compilation} also employs the
nanopass approach, breaking up the compiler into many small passes
that are easier for student to digest~\citep{Sarkar:2004fk}.

This paper presents a proof of correctness in Agda for four nanopasses
that translate a source language with let binding, integer arithmetic,
conditional expressions and Booleans into an x86-flavored register
transfer language. These four passes correspond to the
following three passes in \emph{Essentials of Compilation}:
\begin{enumerate}
\item Remove Complex Operands
\item Explicate Control
\item Select Instructions
\end{enumerate}
The process of proving of correctness of the Explicate Control pass
made it clear to me that it has two distinct reponsibilities that
deserve to be separated:
\begin{itemize}
\item Replacing \texttt{let} expressions with assignment statements.
\item Translating Boolean expressions and conditionals into
  \texttt{goto}-based control-flow.
\end{itemize}
(The RTLgen pass of CompCert also handles the above two
responsibilities.)  So in this paper, the Explicate Control pass is
split into two passes, the first named Lift Locals and the second that
keeps the name Explicate Control.
\begin{enumerate}
\item Remove Complex Operands
\item Lift Locals
\item Explicate Control
\item Select Instructions
\end{enumerate}

TODO: roadmap

%% differs in that I also use a big-step semantics for the register
%% transfer language (instead of small step), which streamlines some of
%% the reasoning.

\section{The Languages}

@@

One of the challenges of formalizing semantics in a proof assistant is
that one needs to choose a representation of variables. There are many
options, and I've tried most of them. But just like a game of
whac-a-mole, the savings acheived in one place are lost in others.  My
current favorite are good old de Bruijn indices. The following type
\lstinline{Id} is for identifiers.

@@

@@

\subsection{The C-like language}

@@

The final intermediate language is a subset of x86 assembly language,
except for the addition of variables. A complete compiler would
include a few more passes to perform register allocation and deal with
idiosyncrosies of x86, such as only allowing at most one instruction
argument to access memory.

@@

We now proceed to discuss each compiler pass and its proof of
correctness.

\section{Pass: Remove Complex Operands}

@@

Most of the cases of \lstinline{rco} are trivial, but let us focus on
the case for subtraction. We shall start with a strawman
translation. We recursively compile the subexpressions $e_1$ and
$e_2$. The results might be complex (not atomic), so we
\lstinline{let} binding them both and replace them in the subtraction
with variables (de Bruijn indices 0 and 1). Note that we must shift
the variables in $e_2$ up by one because we inserted a new
\lstinline{let} above $e_2$. (In general, \lstinline{shift-mon m c}
increases by one each variable in $m$ that is greater or equal to the
cutoff $c$.)

@@

\subsection{Remove Complex Operands is Correct}

@@

\section{Pass: Lift Locals}

@@

The variables in \lstinline{e$₂$} also need to be shifted.  They are
shifted up by $i$ because we are moving all of the $i$
\lstinline{let}'s for \lstinline{e$₁$} above
\lstinline{e$₂$}. Regarding the variable that serves as the left-hand
side of the assignment, its de Bruijn index is $i + j$ because the
original index for this \lstinline{Let} was $0$ but we have moved $i +
j$ other \lstinline{let}s above this variable occurence.

@@

\subsection{Lift Locals is Correct}

@@

This theorem is a corollary of the following main lemma, which is
proved by structural induction on the expression \lstinline{m}.
Considerable care is needed in the statement of this lemma so that the
induction hypotheses can be used. Recall that when interpreting an
expression, the environment $ρ$ provides the meaning for the
\emph{free variables} in the expression.  As the interpreter descends
under \lstinline{let} bindings, it pushes values on the front of the
environment. However, after Lift Locals, all of the variables are
bound at the top, so the initial environment won't be empty, but
instead will have entries for all of the variables (we arbitrarily
initialize them to $0$). So in general, we can separate the
environment into two halves, $ρ₁$ and $ρ₂$, where $ρ₂$ is for the free
variables in \lstinline{m} whereas \lstinline{$ρ₁$ ++ $ρ₂$} is for the
free variables in the result expressoin \lstinline{e} of
\lstinline{lift-locals-mon m}. The length of $ρ₁$ corresponds to the
number returned by \lstinline{lift-locals-mon m}. Also, note that
because \lstinline{Mon} is a pure language, \lstinline{e} contains a
single assignment to each variable, so $ρ₂$ is unchanged by the
execution of \lstinline{e}. ($ρ₁$ is changed as each of its variables
are assigned to for the first time.)

@@

\section{Pass: Explicate Control}

@@

We implement the Explicate Control pass with four functions in Agda,
the \lstinline{explicate} function translates a whole program from the
imperative IR to the C-like language, but it delegates all the work to
the other three functions. These three functions are organized
according to the context in which they are called. Reviewing the
grammar of \lstinline{Imp-Exp} (section~\ref{sec:imp}), there are
three contexts in which an \lstinline{Imp-Exp} may appear:
\begin{enumerate}
\item Tail position: the result of the expression is the output of the program,
\item Assignment: the result of the expression is assigned to a variable, and
\item Predicate: the (Boolean) result of the expression determines which
    branch to take in a conditional.
\end{enumerate}

@@

\subsection{Explicate Control is Correct}

@@

The case for conditional expressions also two lemmas about the
monotonicity of the control flow graph.  The induction hypothesis for
\lstinline{e$₂$} says that the generated \lstinline{c$₂$} evaluates
using the CFG that was produced by the compilation of
\lstinline{e$₂$}, but to satisfy the induction hypothesis
for \lstinline{e$₁$} we need \lstinline{c$₂$} evaluated using the CFG
that was produced by the compilation of \lstinline{e$₃$}.
We define the following notation for
expression that one CFG is a prefix of another CFG:
\begin{lstlisting}
G$₁$ $↝$ G$₂$ = $∃$[ G ] G$₁$ ++ G $≡$ G$₂$
\end{lstlisting}
\noindent We prove that the ouput CFG of \lstinline{explicate-tail} is
an extension of its input CFG.
\begin{lstlisting}
explicate-tail-CFG : $∀${e}{G$₁$ G$₂$}{c}
  $→$ explicate-tail e G$₁$ $≡$ (c , G$₂$)
  $→$ G$₁$ $↝$ G$₂$
\end{lstlisting}
\noindent And we prove that the big-step semantics is monotonic with
respect to the $↝$ relation on control flow graphs.
\begin{lstlisting}
eval-CFG : $∀${c}{$ρ$ $ρ′$}{B$₁$ B$₂$}{s s$′$}{v}
  → B$₁$ $↝$ B$₂$
  → (s , $ρ$) , B$₁$ $⊢ᶜ$ c $⇓$ v $⊣$ (s$′$ , $ρ′$)
  → (s , $ρ$) , B$₂$ $⊢ᶜ$ c $⇓$ v $⊣$ (s$′$ , $ρ′$)
\end{lstlisting}

@@

\noindent The case for conditional expressions needs another lemma
regarding control flow graphs. Similar to \lstinline{explicate-tail},
we need a lemma which states that the ouput CFG of
\lstinline{explicate-assign} is an extension of its input CFG.

@@

\section{Pass: Select Instructions}

\section{The Compiler is Correct}

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{all}

@@

>>>>

@@

\begin{abstract}
We present a proof of correctness in Agda for four nanopasses that
translate a source language with let binding, integer arithmetic,
conditional expressions and Booleans into an x86-flavored register
transfer language. The most interesting of these four nanopasses is a
translation of conditional expressions into goto-style control flow
that uses the continuation-oriented approach of Olivier Danvy's
one-pass transformation into monadic normal form (2003).
\end{abstract}

@@

The compilation of Boolean expressions and conditionals is a classic
topic in the compiler literature, going back to an algorithm of
\citet{Aho:1986qf} (presented in the form of an attribute grammar)
that translates Boolean expressions and conditionals into gotos. The
compilation of Boolean expressions and conditionals often occurs at an
interesting stage in the compilation pipeline, where the intermediate
language changes from an abstract syntax \emph{tree} into a control
flow \emph{graph}, or equivalently, an intermediate language with
\texttt{goto}. Further, when targeting an assembly language such as
Intel's x86, the naive approach to compiling boolean expressions and
conditionals produces embarrassingly inefficient code. Similar issues
arise in the compilation of languages such as Haskell, where the
case-of-case transformation is employed to optimize dispatching on
algebraic datatypes~\citep{PEYTONJONES19983}.

@@

My first exposure to these ideas were in Kent Dybvig's compiler
course~\citep{Dybvig:2010aa} where he taught the techniques used in
the Chez Scheme compiler~\citep{Dybvig:2006aa}.  One of those
techniques is called destination-driven code
generation~\citep{Dybvig:1990aa}, which adds an extra parameter to the
compilation function to indicate how the context will use the result
value of an expression. The context may (1) ignore the result (and
only care about the expression's side effects), (2) branch based on
interpreting its result as a boolean, (3) assign the result to a
variable, or (4) output the result. These options are encoded as a
disjoint union.  Relating this to Olivier's transformation functions,
the $\mathcal{E}$ function is for contexts that need the actual value
and the $\mathcal{B}$ function is for contexts that branch on the
expression's result. Of course, a pair of functions is isomorphic to a
function on a disjoint union.
\[
(A → C) × (B → C) ≅ (A + B) → C
\]

@@

Towards developing a course on verified compilation, I am interested
in adapting verification techniques from CompCert to the compiler
presented in the \emph{Essentials of Compilation}
textbook~\citep{Siek:2023tr,Siek:2023ue}. That book takes an
incremental approach to teaching compilers, that is, it guides the
students through the creation of multiple compilers for a sequence of
languages, each of which adds a language feature to the previous
language. For the language with Boolean expressions and conditionals,
the Explicate Control pass of the compiler uses the one-function per
context approach of Danvy to translate into an x86-flavored register
transfer language.  \emph{Essentials of Compilation} also employs the
nanopass approach, breaking up the compiler into many small passes
that are easier for student to digest~\citep{Sarkar:2004fk}.

This paper presents a proof of correctness in Agda for four nanopasses
that translate a source language with let binding, integer arithmetic,
conditional expressions and Booleans into an x86-flavored register
transfer language. These four passes correspond to the
following three passes in \emph{Essentials of Compilation}:
\begin{enumerate}
\item Remove Complex Operands
\item Explicate Control
\item Select Instructions
\end{enumerate}
The process of proving of correctness of the Explicate Control pass
made it clear to me that it has two distinct responsibilities that
deserve to be separated:
\begin{itemize}
\item Replacing \texttt{let} expressions with assignment statements.
\item Translating Boolean expressions and conditionals into
  \texttt{goto}-based control-flow.
\end{itemize}
(The RTLgen pass of CompCert also handles the above two
responsibilities.)  So in this paper, the Explicate Control pass is
split into two passes, the first named Lift Locals and the second that
keeps the name Explicate Control.
\begin{enumerate}
\item Remove Complex Operands
\item Lift Locals
\item Explicate Control
\item Select Instructions
\end{enumerate}

Section~\ref{sec:langs} of the paper introduces the intermediate
languages. Section~\ref{sec:compiler} describes the compiler,
detailing each pass and the proof of correctness.

\section{The Languages}
\label{sec:langs}

@@

One of the perennial challenges of formalizing semantics in a proof
assistant is that one needs to choose a representation of variables.
%% There are many
%% options, and I've tried most of them. But just like a game of
%% whac-a-mole, the savings achieved in one place are lost in others.
Of the many options, my current favorite is good old de Bruijn
indices~\citep{Bruijn:1972kx}.  Thus, identifiers \lstinline{Id} are
represented as natural numbers.

@@

(This intermediate language does not correspond to the monadic normal
form in Olivier's paper. It is the language of section
\ref{sec:c-lang} that corresponds to the target of Olivier's one-pass
transformation.)

@@

\subsection{The C-like language}
\label{sec:c-lang}

@@

The final intermediate language is a subset of x86 assembly language,
except for the addition of variables. A complete compiler would
include a few more passes to perform register allocation and deal with
idiosyncrasies of x86, such as only allowing at most one instruction
argument to access memory.

@@

We now proceed to discuss the compiler and its proof of correctness.

\section{The Compiler}
\label{sec:compiler}

The compiler is simply the composition of the four passes,
which we detail in the next four subsections.

\begin{lstlisting}
compile e = (select-inst ∘ explicate ∘ lift-locals ∘ rco) e
\end{lstlisting}

\subsection{Pass: Remove Complex Operands}

@@

Most of the cases of \lstinline{rco} are trivial, but let us focus on
the case for subtraction. We shall start with a straw-man
translation. We recursively compile the subexpressions $e_1$ and
$e_2$. The results might be complex (not atomic), so we
\lstinline{let} binding them both and replace them in the subtraction
with variables (de Bruijn indices 0 and 1). Note that we must shift
the variables in $e_2$ up by one because we inserted a new
\lstinline{let} above $e_2$. (In general, \lstinline{shift-mon m c}
increases by one each variable in $m$ that is greater or equal to the
cutoff $c$.)

@@

\subsubsection{Remove Complex Operands is Correct}

@@

\subsection{Pass: Lift Locals}

@@

The variables in \lstinline{e$₂$} also need to be shifted.  They are
shifted up by $i$ because we are moving all of the $i$
\lstinline{let}'s for \lstinline{e$₁$} above
\lstinline{e$₂$}. Regarding the variable that serves as the left-hand
side of the assignment, its de Bruijn index is $i + j$ because the
original index for this \lstinline{Let} was $0$ but we have moved $i +
j$ other \lstinline{let}s above this variable occurrence.

@@

\subsubsection{Lift Locals is Correct}

@@

This theorem is a corollary of the following main lemma, which is
proved by structural induction on the expression \lstinline{m}.
Considerable care is needed in the statement of this lemma so that the
induction hypotheses can be used. Recall that when interpreting an
expression, the environment $ρ$ provides the meaning for the
\emph{free variables} in the expression.  As the interpreter descends
under \lstinline{let} bindings, it pushes values on the front of the
environment. However, after Lift Locals, all of the variables are
bound at the top, so the initial environment won't be empty, but
instead will have entries for all of the variables (we arbitrarily
initialize them to $0$). So in general, we can separate the
environment into two halves, $ρ₁$ and $ρ₂$, where $ρ₂$ is for the free
variables in \lstinline{m} whereas \lstinline{$ρ₁$ ++ $ρ₂$} is for the
free variables in the result expression \lstinline{e} of
\lstinline{lift-locals-mon m}. The length of $ρ₁$ corresponds to the
number returned by \lstinline{lift-locals-mon m}. Also, note that
because \lstinline{Mon} is a pure language, \lstinline{e} contains a
single assignment to each variable, so $ρ₂$ is unchanged by the
execution of \lstinline{e}. ($ρ₁$ is changed as each of its variables
are assigned to for the first time.)

@@

\subsection{Pass: Explicate Control}

@@

We implement the Explicate Control pass with four functions in Agda,
the \lstinline{explicate} function translates a whole program from the
imperative IR to the C-like language, but it delegates all the work to
the other three functions. These three functions are organized
according to the context in which they are called. Reviewing the
grammar of \lstinline{Imp-Exp} (subsection~\ref{sec:imp}), there are
three contexts in which an \lstinline{Imp-Exp} may appear:
\begin{enumerate}
\item Tail position: the result of the expression is the output of the program,
\item Assignment: the result of the expression is assigned to a variable, and
\item Predicate: the (Boolean) result of the expression determines which
    branch to take in a conditional.
\end{enumerate}

@@

\subsubsection{Explicate Control is Correct}

@@

The case for conditional expressions also two lemmas about the
monotonicity of the control flow graph.  The induction hypothesis for
\lstinline{e$₂$} says that the generated \lstinline{c$₂$} evaluates
using the CFG that was produced by the compilation of
\lstinline{e$₂$}, but to satisfy the induction hypothesis
for \lstinline{e$₁$} we need \lstinline{c$₂$} evaluated using the CFG
that was produced by the compilation of \lstinline{e$₃$}.
We define the following notation for
expression that one CFG is a prefix of another CFG:
\begin{lstlisting}
G$₁$ $↝$ G$₂$ = $∃$[ G ] G$₁$ ++ G $≡$ G$₂$
\end{lstlisting}
\noindent We prove that the output CFG of \lstinline{explicate-tail} is
an extension of its input CFG.
\begin{lstlisting}
explicate-tail-CFG : $∀${e}{G$₁$ G$₂$}{c}
  $→$ explicate-tail e G$₁$ $≡$ (c , G$₂$)
  $→$ G$₁$ $↝$ G$₂$
\end{lstlisting}
\noindent And we prove that the big-step semantics is monotonic with
respect to the $↝$ relation on control flow graphs.
\begin{lstlisting}
eval-CFG : $∀${c}{$ρ$ $ρ′$}{B$₁$ B$₂$}{s s$′$}{v}
  → B$₁$ $↝$ B$₂$
  → (s , $ρ$) , B$₁$ $⊢ᶜ$ c $⇓$ v $⊣$ (s$′$ , $ρ′$)
  → (s , $ρ$) , B$₂$ $⊢ᶜ$ c $⇓$ v $⊣$ (s$′$ , $ρ′$)
\end{lstlisting}

@@

\noindent The case for conditional expressions needs another lemma
regarding control flow graphs. Similar to \lstinline{explicate-tail},
we need a lemma which states that the output CFG of
\lstinline{explicate-assign} is an extension of its input CFG.

@@

\subsection{Pass: Select Instructions}

The responsibility of the Select Instructions pass is to convert the
assignment statements and primitive operations of the C-like language
into x86 instructions. The minor challenge is that many x86
instructions take just two arguments and have an update-in-place
semantics.

The function \lstinline{select-inst} translation a whole program in
the C-like language to x86.
\begin{lstlisting}
select-inst (Program n l B) =
  Program n l (map select-stmt B)
\end{lstlisting}

\noindent The \lstinline{select-stmt} function maps each statement to
one or more instructions. The return statement is compiled like an
assignment to the \lstinline{rax} register. (The calling convention
for x86 says that the return value of a function is passed in
\lstinline{rax}.)

\begin{lstlisting}
select-stmt (Return e) = select-assign e (Reg rax)
\end{lstlisting}

\noindent For the assignment statement we delegate to the
\lstinline{select-assign} function, which analyzes the expression
\lstinline{e} to determine which instruction to generate.
We describe \lstinline{select-assign} shortly.

\begin{lstlisting}
select-stmt (Assign x e c) =
  select-assign e (Var x) ++ select-stmt c
\end{lstlisting}

\noindent The \lstinline{IfEq} of the C-like language
is translated to a \lstinline{CmpQ} instruction followed
by \lstinline{JmpEq} and an unconditional \lstinline{Jmp}.

\begin{lstlisting}
select-stmt (IfEq a₁ a₂ thn els) =
  CmpQ (to-arg a₁) (to-arg a₂) ∷
  JmpEq thn ∷
  Jmp els ∷ []
\end{lstlisting}

\noindent Finally, we translate \lstinline{Goto} into a \lstinline{Jmp}.
\begin{lstlisting}
select-stmt (Goto l) = Jmp l ∷ [] 
\end{lstlisting}

Next we look at some representative cases of
\lstinline{select-assign}. The first parameter is the right-hand side
of the assignment and the second parameter is the left-hand side, that
is, the destination. For subtraction, we use the \lstinline{rax}
register as scratch space. We first move argument \lstinline{a$₁$}
into \lstinline{rax}, then subtract \lstinline{a$₂$} from
\lstinline{rax}, and finally move the result into \lstinline{dest}.

\begin{lstlisting}
select-assign (Sub a$₁$ a$₂$) dest =
  MovQ (to-arg a$₁$) (Reg rax) $∷$
  SubQ (to-arg a$₂$) (Reg rax) $∷$
  MovQ (Reg rax) dest $∷$ []
\end{lstlisting}

\noindent Of course there are special cases that require fewer
instructions. For example, if the first argument of the subtraction is
identical to the destination, then we can simply generate one
\lstinline{SubQ} instruction.

\begin{lstlisting}
select-assign (Sub (Var x) a$₂$) (Var x) = 
  SubQ (to-arg a$₂$) (Var x) $∷$ []
\end{lstlisting}

\noindent For the sake of keeping the proof short, we did not include
any of the special cases.

We compile the assignment of an equality comparison into a
\lstinline{CmpQ} and a \lstinline{MovQ}. The \lstinline{CmpQ} always
puts its result into \lstinline{rax}, so the \lstinline{MovQ} is
needed to place the result into the destination.\footnote{The real
\lstinline{cmpq} instruction is slightly stranger, involving
the \lstinline{EFLAGS} register.}

\begin{lstlisting}
select-assign (Eq a$₁$ a$₂$) dest =
  CmpQ (to-arg a$₁$) (to-arg a$₂$) $∷$
  MovQ (Reg rax) dest $∷$ []
\end{lstlisting}

\subsubsection{Select Instructions is Correct}

The correctness theorem for Select Instructions says that if a program
output is $v$ given input $s$, then Explicate Control produces a
program that outputs the same value given the same input.

\begin{lstlisting}
select-inst-correct : $∀${p}{s}{v}
  → interp-prog p s v
  → interp-x86-var (select-inst p) s v
\end{lstlisting}

The correctness theorem delegates to the correctness lemma for
\lstinline{select-stmt}, which says that if statement \lstinline{c}
evaluates to \lstinline{v}, then the instructions generated for
\lstinline{c} put the same value \lstinline{v} into \lstinline{rax}.
The list of x86 instruction blocks is obtain by mapping
\lstinline{select-stmt} over the control-flow graph \lstinline{G}.

\begin{lstlisting}
select-stmt-correct : $∀$ (c : CStmt) ($ρ$ $ρ′$ : Env Value)
  (s s$′$ : Inputs)(regs : List Value)(v : Value)(G : CFG)
  $→$ (s , $ρ$) , G $⊢ᶜ$ c $⇓$ v $⊣$ (s$′$ , $ρ′$)
  $→$ 0 < length regs
  $→$ $∃$[ regs$′$ ] $∃$[ b ] (s , regs , $ρ$) , map select-stmt G
         $⊩$ select-stmt c $⇓$ (s$′$ , regs$′$ , $ρ′$) , b
    × nth regs$′$ rax $≡$ just v
\end{lstlisting}

When the statement is an assignment, the proof uses the following
lemma. It states that if expression \lstinline{e} evaluates to
\lstinline{v}, then the instructions generated for \lstinline{e} put
\lstinline{v} into the specified destination. The predicate
\lstinline{wrote} specifies the update to either the appropriate
variable or register depending on the kind of destination.

\begin{lstlisting}
select-assign-correct : $∀$ (e : CExp) ($ρ$ : Env Value)
   (s s$′$ : Inputs) (dest : Dest) (regs : List Value)
   (v : Value) (B : List Block) 
  $→$ interp-CExp e $ρ$ s $≡$ just (v , s$′$)
  $→$ 0 < length regs
  $→$ DestOK dest 1
  $→$ $∃$[ st$′$ ] (s , regs , $ρ$) , B $⊩$
            select-assign e dest $⇓$ st$′$ , true
      × wrote dest v (s$′$ , regs , $ρ$) st$′$
\end{lstlisting}

\noindent The premise \lstinline{0 < length regs} ensures that we at
least have one register (\lstinline{rax}). The premise
\lstinline{DestOK dest 1} ensures that the destination is either a
variable or the register (\lstinline{rax}).

\subsection{The Compiler is Correct}

The correctness theorem for the compiler composes the four correctness
lemmas of the previous subsections.

\begin{lstlisting}
compile-correct : (e : Exp) (s : Inputs) (v : Value)
  $→$ interp-LIf e s $≡$ just v
  $→$ interp-x86-var (compile e) s v
compile-correct e s v ie =
   let i-rco = trans (rco-correct e s) ie in
   let i-lift = lift-locals-correct (rco e) s v i-rco in
   let i-exp = explicate-correct i-lift in
   select-inst-correct i-exp
\end{lstlisting}

\section{Conclusion}

This paper takes a step in adapting the literature on compiler
correctness~\citep{Leroy:2006fe,Kumar:2014aa} to a university course
based on the incremental and nanopass approach of \emph{Essentials of
Compilation}~\citep{Siek:2023tr,Siek:2023ue}. We prove the correctness
of a compiler for a source language with let binding, integer
arithmetic, conditional expressions and Booleans, targeting an
x86-flavored register transfer language. The process of developing
mechanizing the proof in Agda encouraged a further split in the
nanopasses, separating Explicate Control into two passes, one that
lifts \lstinline{let} bindings and the second that compiles structured
control-flow into \lstinline{goto}. In designing these compiler passes
and proofs, it is enjoyable to trace the history of ideas going back
to \citet{Aho:1986qf}, \citet{Dybvig:1990aa}, and Olivier Danvy's
one-pass transformation into monadic normal form~\citep{Danvy:2003fk}.

%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{all}

@@

% LocalWords:  sigplan tex docstrip dtx bibtex TeXCount TC envir Siek
% LocalWords:  displaymath Nanopasses authornote authornotemark Tobin
% LocalWords:  rv Trovato al Agda nanopasses Olivier Danvy's ccs desc
% LocalWords:  nanopass Ichiro Aho qf gotos datatypes PEYTONJONES fk
% LocalWords:  Danvy arity Dybvig's Dybvig aa Chez Olivier's CompCert
% LocalWords:  fe Coq RTLgen CminorSel CompCert's CakeML stackLang tr
% LocalWords:  labLang Kumar ue Sarkar whac de Bruijn kx Eq Exp Num
% LocalWords:  Bool Var Int Env interp exp LIf immediates Atm atm mon
% LocalWords:  LMonIf IR AST Prog StateImp cmpq je jmp CExp CStmt CFG
% LocalWords:  IfEq CProg prog Arg Reg Dest Inst MovQ SubQ CmpQ JmpEq
% LocalWords:  ReadInt rax StateX EFLAGS var inst rco cmplx suc IL ie
% LocalWords:  retB thenB pred GHC thn eval Agda's transl condexpr
% LocalWords:  stmt arg dest regs DestOK trans
